---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# methylCIPHERv2

<!-- badges: start -->
<!-- badges: end -->

`{methylCIPHERv2}` scores CpG-based DNA methylation ages, commonly called epigenetic clocks. It reads a beta matrix and returns one score per sample per clock, together with the coverage counts that say how much of each clock's CpG panel was present in the data.

`calc_clocks()` is the single entry point. Every other function either describes the available clocks before the calculation, or inspects the results after it.

## Installation

`{methylCIPHERv2}` is not on CRAN yet.

You can install the development version of `{methylCIPHERv2}` with:

```r
pak::pkg_install("hhp94/methylCIPHERv2")
```

## Workflow

```{r setup}
library(methylCIPHERv2)
```

### Browse the clocks

`list_clocks()` returns a data.frame showing all the supported clocks.

```{r}
head(list_clocks())
```

A clock can be selected by `clock_id`, `group_id`, or `tag`. `list_clock_tags()` returns every tag a `clocks` argument accepts.

```{r}
list_clock_tags()
```

`clock_cpgs()` returns the full CpG panel behind any set of clocks.

```{r}
grim_age_cpgs <- clock_cpgs("GrimAge")
head(grim_age_cpgs)
length(grim_age_cpgs)
```

### Download the external weights

Most clocks ship inside the package. A few require downloadable assets. `list_mc_assets()` reports each group, its size, and whether it is already on disk.

```{r, include = FALSE}
old <- set_mc_assets_dir(tempdir())
```

```{r}
list_mc_assets()
```

For example, `"SystemsAge"` needs its asset. `download_mc_assets()` downloads these assets.

```{r, eval = FALSE}
# pass "SystemsAge" to download that group alone
download_mc_assets(groups = "all", ask = TRUE) # downloading requires confirmation
```

```{r, include = FALSE}
set_mc_assets_dir(old)
```

```{r}
list_mc_assets()
```

The function `clear_mc_assets()` delete the downloaded assets with confirmation. `vignette("assets")` covers the assets directory and the advanced path settings.

### Calculate Epigenetic Clocks

We simulate some data with the two kinds of missing data: CpGs that are absent altogether, and CpGs that are present but missing in some samples.

```{r}
set.seed(1)
sim <- sim_DNAm(
  "Horvath1",
  n = 6,
  Age = TRUE,
  remove = 10 # drop CpGs entirely, to simulate completely missing CpGs
)

# simulate partially missing CpGs
sim[["DNAm"]][sample.int(length(sim[["DNAm"]]), 20)] <- NA

sim

sum(is.na(sim[["DNAm"]]))
```

The two kinds of missingness are handled differently. A CpG that is absent altogether is either filled from a reference value that ships with the clock, or dropped. The `policy` column of `clocks_coverage()` says which one a clock uses. A CpG that is present but missing in some samples is imputed from the samples that do have it.

Some clocks need a `Female` covariate, which the `covariates` column of `list_clocks()` marks. `predict_sex()` scores the two `DNAmSex_Wang` scores and returns a predicted sex for every sample. When `pheno` does carry `Female`, `predict_sex()` also returns `recorded_sex` and `sex_mismatch`, and reports how many samples disagree.

```{r}
sex_sim <- sim_DNAm(
  c("DNAmSex_Wang_ChrX", "DNAmSex_Wang_ChrY"),
  n = 6,
  Female = TRUE
)

predict_sex(sex_sim[["DNAm"]], sex_sim[["pheno"]])
```

The betas are random here, so every predicted sex is arbitrary and the mismatches are expected.

Clock calculation is performed by the `calc_clocks()` function.

```{r}
res <- calc_clocks(
  sim[["DNAm"]],
  "Horvath1",
  pheno = sim[["pheno"]]
)
res
```

`calc_clocks()` can also run in batches, which are combined afterwards.

```{r}
batch_1 <- sim_DNAm(c("Knight"), n = 6, suffix = "1")
batch_2 <- sim_DNAm(c("Knight"), n = 6, suffix = "2")

r1 <- calc_clocks(batch_1[["DNAm"]], "Knight", pheno = batch_1[["pheno"]])
r2 <- calc_clocks(batch_2[["DNAm"]], "Knight", pheno = batch_2[["pheno"]])

rbind(r1, r2)
```

`as.data.frame()` or `as.matrix()` finalize the scores for further analysis. `as.data.frame()` returns long or wide.

```{r}
# long, the default
head(as.data.frame(res))
# wide
head(as.data.frame(res, long = FALSE))
```

### Check the coverage

Many factors affect the validity of a calculated clock. The most important is missing data, both for each clock and for each sample.

`clocks_coverage()` reports missing data by `clock_id`. Its last column, `missing_cpgs`, names every CpG that was missing (dropped here for width).

```{r}
cov <- clocks_coverage(res, all_columns = TRUE)
cov[setdiff(names(cov), grepv("missing_cpgs|norm|role", names(cov)))]
```

`Horvath1` uses the `omit` policy, so the 10 absent CpGs were dropped rather than filled, and the 20 missing values landed on 19 distinct CpGs that were imputed.

`samples_coverage()` covers missingness per sample per clock.

```{r}
head(samples_coverage(res))
```

### Age acceleration

`calc_accel()` regresses each score on chronological age and returns the residual. The value returned by `calc_clocks()` keeps only the id column and the covariates the calculation required, so chronological age is supplied through `data`, joined on the sample id.

```{r}
head(calc_accel(res, data = sim[["pheno"]]))
```

### Check the age associations

`calc_accel()` works sample by sample. `score_associations()` looks at all the samples at once, and compares each clock's observed correlation with age against a reference range that ships with the package. Age is supplied through `age` here, for the same reason it goes through `data` above.

```{r}
score_associations(res, age = sim[["pheno"]][["Age"]])
```

The scores come from random data, so the correlation itself carries no meaning, and six samples make the reference range wide. What matters in real use are the two flags. `outside` marks a clock whose observed correlation falls outside the reference range, and `wrong_sign` marks one that runs in the opposite direction from a reference correlation stronger than 0.3.

### Cite the clocks

`cite_clocks()` returns the reference for every clock in a result, so a methods section can be assembled from the run itself.

```{r}
cite_clocks(res)
```

Or pass the desired `clock_id` directly.

```{r, eval = FALSE}
cite_clocks("Horvath1")
```
