---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# methylCIPHERv2

<!-- badges: start -->
<!-- badges: end -->

`{methylCIPHERv2}` is a lightweight epigenetic clocks R package. The main function, `calc_clocks()`, has 3 important inputs (see `calc_clocks()` docs for more details):

* `DNAm`: A DNAm beta-matrix where samples are in the rows and CpGs are in the columns
* `clocks`: A list of clocks (e.g., Horvath1), clock "group" (e.g., GrimAge), or tags ("mitotic").
* `pheno`: A data.frame that contains Age and Female (1/0), and an ID column that links to the sample names of the DNAm beta matrix.

## Installation

You can install the development version of `{methylCIPHERv2}` with:

```r
pak::pkg_install("hhp94/methylCIPHERv2")
```

## Workflow

```{r setup}
library(methylCIPHERv2)
```

### Browse the clocks

`list_clocks()` returns a data.frame showing all the supported clocks.

```{r}
head(list_clocks(), n = 3)
```

A clock can be selected by `clock_id`, `group_id`, or `tag`. `list_clock_tags()` returns every tag a `clocks` argument accepts. `clock_cpgs()` returns the full CpG panel behind any set of clocks.

```{r}
grim_age_cpgs <- clock_cpgs("GrimAge")
head(grim_age_cpgs, n = 3)
length(grim_age_cpgs)
```

### Download the external weights

A few clocks require downloadable assets. `list_mc_assets()` reports the required download status.

```{r, include = FALSE}
old <- set_mc_assets_dir(tempdir())
```

```{r}
list_mc_assets()
```

For example, `"SystemsAge"` needs its asset for `calc_clocks()`. `download_mc_assets()` downloads these assets onto your computer.

```{r, eval = FALSE}
# pass "SystemsAge" to download that group alone
download_mc_assets(groups = "all", ask = TRUE) # downloading requires confirmation
```

```{r, include = FALSE}
set_mc_assets_dir(old)
```

```{r}
list_mc_assets()
```

The function `clear_mc_assets()` deletes the downloaded assets. `vignette("assets")` covers the assets directory and the advanced path settings.

### Calculate Epigenetic Clocks

We simulate some data that carries the two kinds of missing data: CpGs that are absent altogether, and CpGs that are present but missing in some samples.

```{r}
set.seed(1)
sim <- sim_DNAm(
  c("Horvath1", "Bohlin96"),
  n = 6,
  Age = TRUE,
  remove = 10 # drop CpGs entirely, to simulate completely missing CpGs
)

# simulate partially missing CpGs
sim[["DNAm"]][sample.int(length(sim[["DNAm"]]), 20)] <- NA

# count partial NAs
sum(is.na(sim[["DNAm"]]))
```

The two kinds of missingness are handled differently. A CpG that is absent altogether is either filled from a reference value that ships with the clock, or dropped. A CpG that is present but missing in some samples is column mean imputed (i.e., CpG-wise).

Some clocks need a `Female` covariate. If not readily available, `predict_sex()` scores the two `DNAmSex_Wang` ChrX and ChrY clocks and returns a predicted sex for every sample. When `pheno` provides `Female`, `predict_sex()` also returns `recorded_sex` and `sex_mismatch`, and reports how many samples disagree.

```{r, eval = F}
# The betas are random here, so the mismatches are expected.
sex_sim <- sim_DNAm(
  c("DNAmSex_Wang_ChrX", "DNAmSex_Wang_ChrY"),
  n = 6,
  Female = TRUE
)

predict_sex(sex_sim[["DNAm"]], sex_sim[["pheno"]])
#>        ID DNAmSex_Wang_ChrX DNAmSex_Wang_ChrY predicted_sex recorded_sex
#> 1 sample1          37.76046         -4.768303        Female       Female
#> 2 sample2          35.86986         -4.877966        Female       Female
```

Clock calculation is performed by the `calc_clocks()` function.

```{r}
res <- calc_clocks(
  DNAm = sim[["DNAm"]],
  clocks = c("Horvath1", "Bohlin96"),
  pheno = sim[["pheno"]]
)
```

`calc_clocks()` can also run in batches and be combined afterwards.

```{r, eval = F}
batch_1 <- sim_DNAm(c("Knight"), n = 6, suffix = "1")
batch_2 <- sim_DNAm(c("Knight"), n = 6, suffix = "2")

r1 <- calc_clocks(batch_1[["DNAm"]], "Knight", pheno = batch_1[["pheno"]])
r2 <- calc_clocks(batch_2[["DNAm"]], "Knight", pheno = batch_2[["pheno"]])

rbind(r1, r2)
```

`as.data.frame()` or `as.matrix()` finalizes the scores for further analysis. `as.data.frame()` returns long or wide.

```{r}
# long, the default
head(as.data.frame(res), n = 3)
# wide
head(as.data.frame(res, long = FALSE), n = 3)
```

### Check the coverage

Many factors affect the validity of a calculated clock. One of the most important is missing data, both for each clock and for each sample.

`clocks_coverage()` reports missing data by `clock_id`. Its last column, `missing_cpgs`, names every CpG that was missing (dropped here for width).

```{r}
cov <- clocks_coverage(res, all_columns = TRUE)
h1 <- cov[cov[["clock_id"]] == "Horvath1", ]
cov[setdiff(names(cov), grep("missing_cpgs|norm|role", names(cov), value = TRUE))]
```

`Horvath1` uses the `omit` policy, so the `r h1[["score_dropped"]]` absent CpGs were dropped rather than filled, and the 20 missing values landed on `r h1[["score_imputed_partial"]]` distinct CpGs that were imputed.

`samples_coverage()` covers missingness per sample per clock.

```{r}
head(samples_coverage(res), n = 3)
```

### Age acceleration

`calc_accel()` regresses each score on chronological age and returns the residual (`type = "accel"`) or calculates the raw difference between a clock and chronological age (`type = "diff"`). The value returned by `calc_clocks()` keeps only the id column and the covariates required for the calculation, so chronological age is supplied through `data` here.

```{r}
# long, the default
head(calc_accel(res, data = sim[["pheno"]]), n = 3)
# can also return wide
head(calc_accel(res, data = sim[["pheno"]], long = FALSE), n = 3)
```

### Check the age associations

`score_associations()` compares each clock's observed correlation with age against a reference range that ships with the package. Age is supplied through `age` here, for the same reason it goes through `data` above.

```{r}
# Random data for illustration purposes only
score_associations(res, age = sim[["pheno"]][["Age"]])
```

### Cite the clocks

`cite_clocks()` returns the reference for every clock in a result.

```{r, eval = F}
cite_clocks(res)
```

Or pass the desired `clock_id` directly.

```{r, eval = FALSE}
cite_clocks("Horvath1")
```
