---
title: "Managing Clock Assets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing Clock Assets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Most clocks ship inside the package and need no setup. Four clock groups are
different. Their weights are too large to ship, so they live in downloadable
files called assets.

This guide shows how to find the directory that holds the assets, how to fill
it, and how to empty it. Every code block that downloads or deletes a file is
shown but not run.

```{r setup}
library(methylCIPHERv2)
```

## The clocks that need an asset

The `external` column of `list_clocks()` marks a clock whose weights are an
asset.

```{r}
catalog <- list_clocks()
unique(catalog[["group_id"]][catalog[["external"]]])
```

A request for any other clock needs no download.

## Where the assets go

`get_mc_assets_dir()` reports the directory in effect.

```{r, eval = FALSE}
get_mc_assets_dir()
#> [1] "/home/you/.cache/R/methylCIPHERv2"
```

Four sources name that directory. The first one that is set wins.

1. The `ext_data` argument of `calc_clocks()`, `clock_cpgs()` and
   `load_mc_assets()`.
2. The `mc.assets_dir` option, which `set_mc_assets_dir()` writes.
3. The `MC_ASSETS_DIR` environment variable.
4. The default cache directory, from `tools::R_user_dir()`.

The default is a cache directory, so the operating system may reclaim it. Set
one of the other three for a directory you control.

## Seeing what is on disk

`list_mc_assets()` answers every question about the directory without changing
it. It never prompts, never downloads, and never deletes.

```{r, eval = FALSE}
list_mc_assets()
#>    group_id n_clocks n_cpgs      size downloaded superseded superseded_size
#> 1 PCBrainAge        1  78464   61.2M       TRUE          0              0B
#> 2   PCClocks        8  78464  240.1M      FALSE          0              0B
#> 3 SystemsAge       12  78464   88.4M       TRUE          1           87.9M
#> 4  Zhang2019        1 319607   12.4M      FALSE          0              0B
```

`downloaded` says whether the current asset is on disk. `superseded` counts
older assets for that group, which a later version of the package replaced.
Superseded assets waste space and nothing reads them.

## Downloading an asset

`download_mc_assets()` writes the assets to disk. It shows the size first and
asks for consent.

```{r, eval = FALSE}
download_mc_assets("SystemsAge")
```

Pass `"all"` for every group, or a vector of group names for some of them.

```{r, eval = FALSE}
download_mc_assets(c("SystemsAge", "PCClocks"))
```

`calc_clocks()` downloads a missing asset for you, under the same prompt. A
separate download is useful to prepare a machine before an analysis starts.

## Loading an asset into memory

`load_mc_assets()` reads the assets into memory and returns them. Pass the
result to `ext_data` to score without touching the disk again.

```{r, eval = FALSE}
assets <- load_mc_assets("SystemsAge")
res <- calc_clocks(DNAm, "SystemsAge", ext_data = assets)
```

This is the fastest option for repeated scoring in one session.

## Choosing another directory

`set_mc_assets_dir()` points the session at another directory. It creates the
directory if it does not exist yet.

```{r, eval = FALSE}
old <- set_mc_assets_dir("~/methylation-assets")
```

The return value restores the previous state, in the manner of `setwd()`. It
is the override that was in place, and it is `NULL` when no override was set.
Pass it back to undo the change.

```{r, eval = FALSE}
set_mc_assets_dir(old)
```

The setting lasts for the session only. For a permanent choice, add the
directory to your `.Renviron` file.

```
MC_ASSETS_DIR=/data/methylation-assets
```

## A directory that never downloads

The `ext_data` argument takes a path. A path restricts the run to the assets
already in that directory. Nothing is downloaded, and a missing asset is an
error rather than a prompt.

```{r, eval = FALSE}
res <- calc_clocks(DNAm, "SystemsAge", ext_data = "/data/methylation-assets")
```

Use this form on a machine with no network access, or in a pipeline that must
not reach out to the internet. Leave `ext_data` at its default of `NULL` to
allow a download.

## Deleting the assets

`clear_mc_assets()` empties the directory. It lists the files and asks for
consent first.

```{r, eval = FALSE}
clear_mc_assets()
```

The delete covers the current assets and the superseded ones together. Pass a
group name to delete one group.

```{r, eval = FALSE}
clear_mc_assets("SystemsAge")
```

Nothing else in the directory is touched. A file the package did not write
stays where it is.

## Working without a prompt

`download_mc_assets()`, `load_mc_assets()` and `clear_mc_assets()` all refuse
to run in a non-interactive session, because nobody can answer the prompt.
Pass `ask = FALSE` to give consent ahead of time.

```{r, eval = FALSE}
download_mc_assets("SystemsAge", ask = FALSE)
```

Use `ask = FALSE` in a script, a scheduled job, or a container build. Consider
the size of the download before you do.
