---
title: "Managing Clock Assets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing Clock Assets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Most clocks are included with the package and need no setup. Some clock groups are too large, so their weights live in downloadable files (i.e., assets).

This vignette shows how `{methylCIPHERv2}` resolves the assets directory, and how to change it.

```{r setup}
library(methylCIPHERv2)
```

## Discover

`list_mc_assets()` shows every clock group that needs a downloaded asset.

```{r, include = FALSE}
old <- set_mc_assets_dir(tempdir())
```

```{r}
list_mc_assets()
```

```{r, include = FALSE}
set_mc_assets_dir(old)
```

`downloaded` says whether the current asset is on disk. `superseded` counts older assets for that group, which a later version of the package replaced. Superseded assets waste space and nothing reads them.

## Pathing

`get_mc_assets_dir()` returns the path in effect.

```{r, eval = FALSE}
get_mc_assets_dir()
#> "C:/Users/<you>/AppData/Local/R/cache/R/methylCIPHERv2"
```

To override that path for this session, run `set_mc_assets_dir()`, which returns the previous setting. The same function restores the default. This is useful for an HPC workflow, where the path can point at a folder of assets you uploaded ahead of time.

```{r, eval = FALSE}
old <- set_mc_assets_dir("/your/override")

# restore the previous setting
set_mc_assets_dir(old)

# or clear the override and fall back to the default
set_mc_assets_dir(NULL)
```

## Durable override

The default path is a cache directory, which a cleanup tool or the operating system can wipe (more likely on macOS and Linux). For a durable path, set the `MC_ASSETS_DIR` environment variable in a user or project `.Renviron` file.

```{r, eval = FALSE}
# open the user .Renviron file
file.edit("~/.Renviron")
```

Add this line to it, then restart R.

```
MC_ASSETS_DIR=/your/override
```

## Downloading

`download_mc_assets()` writes the assets to disk, and always asks for confirmation. Use `ask = FALSE` in a script, a scheduled job, or a container build. Consider the size of the download before you do.

```{r, eval = FALSE}
download_mc_assets("SystemsAge", ask = TRUE) # the default, prompts first

# calc_clocks() routes through the same mechanism. ask = FALSE
# authorizes a download when one is needed.
calc_clocks(DNAm, clocks = "SystemsAge", ask = FALSE)
```

Pass `"all"` for every group, or a vector of group names for some of them.

```{r, eval = FALSE}
download_mc_assets(c("SystemsAge", "PCClocks"))
```

## Loading an asset into memory

Assets can be pre-loaded into memory once with `load_mc_assets()`, then re-used for the whole session. Pass the loaded object to the `ext_data` argument of `calc_clocks()`. This reduces the overhead of a repeated scoring workflow.

```{r, eval = FALSE}
assets <- load_mc_assets("SystemsAge")
res <- calc_clocks(DNAm, "SystemsAge", ext_data = assets)
```

## A directory that never downloads

The `ext_data` argument also takes a path. A path restricts the run to the assets already in that directory. Nothing is downloaded, and a missing asset is an error rather than a prompt.

```{r, eval = FALSE}
res <- calc_clocks(DNAm, "SystemsAge", ext_data = "/data/methylation-assets")
```

Use this form on a machine with no network access, or in a pipeline that must not reach the internet. Leave `ext_data` at its default of `NULL` to allow a download.

## Deleting the assets

`clear_mc_assets()` empties the directory. It lists the files and asks for confirmation, and it takes `ask` like the two verbs above.

```{r, eval = FALSE}
clear_mc_assets()
```

The delete covers the current assets and the superseded ones together. Pass a group name to delete one group.

```{r, eval = FALSE}
clear_mc_assets("SystemsAge")
```

A file the package did not write is never touched.
