---
title: "The Clock Catalog"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, include = FALSE}
library(methylCIPHERv2)
```

```{r, echo = FALSE}
catalog <- list_clocks(all_columns = TRUE)

# a sex-routed family is scored through its own name, so its internal members
# are dropped here. They are the rows whose request_as points elsewhere.
catalog <- catalog[catalog[["clock_id"]] == catalog[["request_as"]], ]

reactable::reactable(
  catalog[c("clock_id", "group_id", "covariates", "normalize")],
  searchable = TRUE,
  filterable = TRUE,
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  defaultPageSize = 25,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(25, 50, 100),
  defaultSorted = "group_id",
  columns = list(
    clock_id = reactable::colDef(name = "Clock", minWidth = 200),
    group_id = reactable::colDef(name = "Group", minWidth = 160),
    covariates = reactable::colDef(name = "Covariates", minWidth = 110),
    normalize = reactable::colDef(name = "Normalize", minWidth = 110)
  )
)
```

## What the columns mean

**Clock** is the identifier for a single scored clock. **Group** is the family it belongs to.

**Covariates** lists the columns that `pheno` must carry for that clock. An empty cell means the clock reads methylation values alone.

**Normalize** names the background normalization scheme a clock uses, and an empty cell means it uses none. A `quantile` cell is part of the clock definition and always runs. A `bmiq` cell is optional, and stays off until the `normalize` argument of `calc_clocks()` turns it on.

## Scoring from the catalog

Any value in the **Clock** column works as the `clocks` argument.

```{r, eval = FALSE}
res <- calc_clocks(DNAm, clocks = "Horvath1", pheno = pheno)
```

A group name scores every member of that family.

```{r, eval = FALSE}
res <- calc_clocks(DNAm, clocks = "GrimAge", pheno = pheno)
```

A tag scores every clock that carries it. `list_clock_tags()` returns the tags with their members.

```{r}
names(list_clock_tags())
```

## Filtering the catalog in code

`list_clocks()` returns this catalog as a data frame, so it is available in a script as well as on this page. It takes a tag directly.

```{r}
mitotic <- list_clocks(tag = "mitotic")
mitotic[["clock_id"]]
```

The result is an ordinary data frame, so you can filter the other columns with base R.

```{r}
all_clocks <- list_clocks()
bundled <- all_clocks[!all_clocks[["external"]], ]
nrow(bundled)
```
